---
title: "BatchQC Examples"
author:
- name: W. Evan Johnson
  affiliation:
  - Division of Infectious Disease, Department of Medicine, Rutgers University 
  - Director, Center for Data Science, Rutgers University
  email: wj183@njms.rutgers.edu 
- name: Jessica McClintock
  affiliation: 
  - Division of Infectious Disease, Department of Medicine, Rutgers University 
  email: jessica.mcclintock@rutgers.edu
package: BatchQC
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{BatchQC Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Example 1: Protein Data
This dataset is from protein expression data captured for 39 proteins.
It has two batches and two conditions corresponding to case and control.

```{r}
library(BatchQC)
data(protein_data)
data(protein_sample_info)
se_object <- BatchQC::summarized_experiment(protein_data, protein_sample_info)
```

# Example 2: Signature Data
This dataset is from signature data captured when activating different growth 
pathway genes in human mammary epithelial cells (GEO accession: GSE73628). 
This data consists of three batches and ten different conditions corresponding 
to control and nine different pathways

```{r}
data(signature_data)
data(batch_indicator)
se_object <- BatchQC::summarized_experiment(signature_data, batch_indicator)
```

# Example 3: Bladderbatch Data
This dataset is from bladder cancer data. This dataset has 57 bladder samples
with 5 batches and 3 covariate levels (cancer, biopsy, control). Batch 1 
contains only cancer, 2 has cancer and controls, 3 has only controls, 4 contains
only biopsy, and 5 contains cancer and biopsy. This data set is from the
bladderbatch package (Leek JT (2023). bladderbatch: Bladder gene expression data
illustrating batch effects. R package version 1.38.0).

```{r}
data(bladder_data)
data(bladder_meta)
se_object <- BatchQC::summarized_experiment(bladder_data, bladder_meta)
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
