---
title: "BatchQC Examples"
author:
- name: W. Evan Johnson
  affiliation:
  - Division of Infectious Disease, Department of Medicine, Rutgers University 
  - Director, Center for Data Science, Rutgers University
  email: wj183@njms.rutgers.edu 
- name: Jessica Anderson
  affiliation: 
  - Division of Infectious Disease, Department of Medicine, Rutgers University 
  email: anderson.jessica@rutgers.edu
- name: Solaiappan Manimaran
  affiliation:
  - Computational Biomedicine, Department of Medicine, Boston University
package: BatchQC
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{BatchQC Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Example 1: Protein Data
This data set is from protein expression data captured for 39 proteins.
It has two batches and two conditions corresponding to case and control.

```{r Protein Sample Data}
library(BatchQC)
data(protein_data)
data(protein_sample_info)
se_object <- BatchQC::summarized_experiment(protein_data, protein_sample_info)
```

# Example 2: Signature Data
This data set is from signature data captured when activating different growth 
pathway genes in human mammary epithelial cells (GEO accession: GSE73628). 
This data consists of three batches and ten different conditions corresponding 
to control and nine different pathways

```{r Signature Sample Data}
data(signature_data)
data(batch_indicator)
se_object <- BatchQC::summarized_experiment(signature_data, batch_indicator)
```

# Example 3: Bladderbatch Data
This data set is from bladder cancer data. This dataset has 57 bladder samples
with 5 batches and 3 covariate levels (cancer, biopsy, control). Batch 1 
contains only cancer, 2 has cancer and controls, 3 has only controls, 4 contains
only biopsy, and 5 contains cancer and biopsy. This data set is from the
bladderbatch package which must be installed to use this data example set (Leek 
JT (2023). bladderbatch: Bladder gene expression data illustrating batch 
effects. R package version 1.38.0).

```{r Bladderbatch Sample Data, eval = FALSE}
if (!requireNamespace("bladderbatch", quietly = TRUE))
    BiocManager::install("bladderbatch")
se_object <- BatchQC::bladder_data_upload()
```

# Example 4: TB Data
This is a whole blood gene expression profiling from well and malnourished
Indian individuals with TB and severely malnourished household contacts with
latent TB infection (LTBI). Severe malnutrition was defined as body mass index
<16. kg/m2 in adults and based on weight-for-height Z scores in children <18
years. Gene expression was measured using RNA-sequencing. (VanValkenburg A, et
al. Malnutrition leads to increased inflammation and expression of tuberculosis
risk signatures in recently exposed household contacts of pulmonary
tuberculosis. Front Immunol. 2022 Sep 28;13:1011166.
doi: 10.3389/fimmu.2022.1011166")

```{r TB Sample Data, eval = FALSE}
if (!requireNamespace("curatedTBData", quietly = TRUE))
    BiocManager::install("curatedTBData")
se_object <- BatchQC::tb_data_upload()
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
